---
title: "Probability"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn concepts relating to probability.
---

```{r setup, include=FALSE, message=FALSE}
library("learnr")
library("gradethis")
# library("gtools")
tutorial_options(exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Basic Concepts

- If you toss a fair coin, the probability that the outcome is a heads is:

```{r p1, exercise=TRUE}

```

```{r p1-check}
grade_result(
  pass_if(~identical(.result, 0.5))
)
```

- If you roll a standard six-sided die, the probability that the outcome is odd is:

```{r p2, exercise=TRUE}

```

```{r p2-check}
grade_result(
  pass_if(~identical(.result, 0.5))
)
```

<div id="p2-hint">
**Hint:** Event $A=$ Outcome is odd. $A=$ \{1,3,5\}. Required probability = $\frac{n(A)}{n(T)}$, where $n(.)$ denotes the number of elements in the set. 
</div>


- Suppose you roll a pair of six-sided dice. Assume that the dice can be distinguished from each other. What is the probability that the sum of the two dice is 8? (rounded up to 2 decimal points)

```{r p2a, exercise=TRUE}

```

```{r p2a-check}
grade_result(
  pass_if(~identical(.result, round(5/36, 2))),
  pass_if(~identical(.result, 5/36))
)
```

<div id="p2-hint">
**Hint:** There are a total of 36 possibilities, out of which the event that the sum of the two dice is 8 can arise in 5 ways - $\{(2,6), (3,5), (4,4), (5,3), (6,2)\}$. Required probability = 5/36. 
</div>


```{r p3-setup, include=FALSE}
pA <- 0.3
pB <- 0.6
pAB <- 0.25
pAorB <- pA + pB - pAB; pAorB
notpAorB <- 1 - pAorB; notpAorB
pAnotB <- pA - pAB; pAnotB
```

- Consider an online shopping portal. Let $A$ denote the event that a shopper has a PayPal account and $B$ the event that the shopper has a Natwest Bank account. Suppose that $P(A) = 0.3$, $P(B) = 0.6$, and $P(A \cap B) = 0.25$.

a. Compute the probability that the individual has at least one of the two types of account.

```{r p3a, exercise=TRUE}

```

```{r p3a-check}
grade_result(
  pass_if(~identical(.result, 0.65))
)
```

<div id="p3a-hint">
**Hint:** $P(A \cup B) = P(A) + P(B) - P(A \cap B)$.
</div>

b. What is the probability that the shopper has neither type of card?
    
```{r p3b, exercise=TRUE}

```

```{r p3b-check}
grade_result(
  pass_if(~identical(.result, 0.35))
)
```

<div id="p3b-hint">
**Hint:** $P(A' \cap B') = P(A \cup B)' = 1 - P(A\cup B)$.
</div>    


c. What is the probability that the shopper has a Paypal account but not a Natwest Bank account?
    
```{r p3c, exercise=TRUE}

```

```{r p3c-check}
grade_result(
  pass_if(~identical(.result, 0.05))
)
```

<div id="p3c-hint">
**Hint:** $P(A \cap B') = P(A) - P(A \cap B)$.
</div> 


- Let $A$ denote the event that a job advertised requires knowledge of the R statistical software, and let $B$ be the event that the next job posted requires knowledge of Stata. Suppose that $P(A) = 0.5$ and $P(B) = 0.4$.

```{r p4-setup, include=FALSE}
pA <- 0.5
pB <- 0.4
pnotA <- 1-pA; pnotA
pAorB <- pA + pB; pAorB
1 - pAorB
```


a. Do the probabilities have to add to 1? Answer "Yes" or "No".

```{r p4a, exercise=TRUE}

```

```{r p4a-check}
grade_result(
  pass_if(~identical(.result, "Yes")),
  pass_if(~identical(.result, "yes")),
)
```

<div id="p4a-hint">
**Hint:** Jobs could require knowledge of other software e.g. Python.
</div> 

b. Calculate $P(A')$

```{r p4b, exercise=TRUE}

```

```{r p4b-check}
grade_result(
  pass_if(~identical(.result, 0.5)),
)
```

<div id="p4b-hint">
**Hint:** $P(A') = 1 - P(A)$
</div> 


c. Calculate $P(A \cup B)$

```{r p4c, exercise=TRUE}

```

```{r p4c-check}
grade_result(
  pass_if(~identical(.result, 0.9)),
)
```

<div id="p4c-hint">
**Hint:** Since $A$ and $B$ are mutually exclusive events, $P(A \cup B) = P(A) + P(B)$
</div> 


d. Calculate $P(A' \cap B')$

```{r p4d, exercise=TRUE}

```

```{r p4d-check}
grade_result(
  pass_if(~identical(.result, 0.1)),
)
```

<div id="p4d-hint">
**Hint:** Draw Venn diagram. Note that $P(A' \cap B') = P(A \cup B)' = 1 - p(A \cup B)$. Holds even if they are not mutually exclusive -- deMorgan's Law.
</div> 


-----

## Counting

- Suppose that three letter are to be selected from A, B, C, D, E and F and arranged in order. How many permutations are possible?

```{r c5, exercise=TRUE}

```

```{r c5-check}
grade_result(
  pass_if(~identical(.result, factorial(6)/factorial(6-3)))
)
```

<div id="c5-hint">
**Hint:** There are 6 letter A to F. Order is important. The first letter can be picked in 6 ways since any of the letters can be picked first. The second can be picked from the remaining 5 in 5 ways and similarly the third in 4 ways. Total number of ways = $6 \times 5 \times 4 = 120$ ways.
</div> 

- An urn contains 10 red, 6 green and 4 yellow balls. Suppose we draw two balls without replacement. What is the probability that the sample contains exactly one red or exactly one yellow ball? (final answer rounded up to 2 dp)

```{r c6-setup, include=FALSE}
# prob (red and green)
p.RG <- (10*6)/choose(20,2); p.RG
# prob green and yellow
p.GY <- (6*4)/choose(20,2); p.GY
# prob red and yellow
p.RY <- (10*4)/choose(20,2); p.RY
# prob exactly one red or exactly one yellow
prob <- p.RG + p.GY + p.RY; round(prob,2)
```


```{r c6, exercise=TRUE}

```


```{r c6-check}
grade_result(
  pass_if(~identical(.result, 0.65))
)
```

<div id="c6-hint">
**Hint:** P(exactly one red or exactly one yellow) = P(red and green) + P(green and yellow) + P(red and yellow).  P(red and green) = $\frac{10 \cdot 6}{\binom{20}{2}}$. In a similar way, calculate the probability for the events P(green and yellow) and P(red and yellow). Add the three probabilities and round up the final result to 2dp.
</div> 

- What is the probability that in a roll of 3 dice no two show the same face? (rounded up to 2 dp)

```{r c7, exercise=TRUE}

```

```{r c7-check}
grade_result(
  pass_if(~identical(.result, 120/216))
)
```

<div id="c7-hint">
**Hint:** Each die has 6 faces. So, total number of outcomes (sample space) is $6^3=216$. The event that no two dice show the same face can happen in $6 \times 5 \times 4 = 120$ ways (first die can take any of the six faces, the second can take any of the remaining five faces, and the third can take any of the remaining 4). Required probability = $120/216=0.56$
</div> 


- Anderson 1.26, pp34


## Birthday problem

This is a computationally challenging problem. Try to work out its logic!

```{r birthday-setup, include=FALSE}
rm(list = ls())

k <- 10
p <- numeric(k)
for (i in 1:k){
  q <- 1 - (0:(i-1))/365
  p[i] <- prod(q)
}
prob <- p[k]
prob
plot(1:k, p)
```

There are 10 people in a room. Assume each person's birthday is equally likely to be any of the 365 days of the year (we exclude leap years) and that birthday's are *independent*. What is the probability that there are no birthday matches (i.e. no two people in the room share the same birthday)?

```{r birthday, exercise=TRUE}

```

```{r birthday-check}
grade_result(
  pass_if(~identical(.result, round((choose(365,10)*factorial(10))/365^10, 2)))
)
```

<div id="birthday-hint">
**Hint:** Any birthday is equally likely. So, there are a total of $365^{10}$ ways. Number of outcomes among them with no repeated birthdays is $P^{365}_{10}$. Probability of no shared birthday = $\frac{P^{365}_{10}}{365^{10}}$. 
</div> 


## Conditional Probability

- A laptop manufacturer found that 75% of all customers want a high-speed processor, 80% want a fingerprint reader, and 65% want both. 

```{r c8-setup, include=FALSE}
# prob high speed processor
pA <- 0.75
# prob fingerprint reader
pB <- 0.80
# prob both
pAB <- 0.65
# prob A or B
pAorB <- pA + pB - pAB; pAorB
pAgB <- pAB/pB; pAgB
pBgA <- pAB/pA; pBgA
```

a. What is the probability that a customer will want at least one of these features? (rounded up to 2 dp)


```{r c8a, exercise=TRUE}

```

```{r c8a-check}
grade_result(
  pass_if(~identical(.result, 0.9))
)
```

<div id="c8a-hint">
**Hint:** $P(A)=$ prob that customer wants high-speed processor. $P(B)=$ customer wants fingerprint reader. By law of total probability, $A(A \text{ or } B) = P(A \cup B) = P(A) + P(B) - P(A \cap B)$.
</div> 


b. What is the probability that a customer who wants a high-speed processor also wants a fingerprint reader? (rounded up to 2 dp)

```{r c8b, exercise=TRUE}

```

```{r c8b-check}
grade_result(
  pass_if(~identical(.result, 0.81))
)
```

<div id="c8b-hint">
**Hint:** From the law of conditional probability: $P(A|B) = \frac{P(A \cap B)}{P(B)}$.
</div> 

c. What is the probability that a customer who wants a fingerprint reader also wants a high-speed processor? (rounded up to 2 dp)

```{r c8c, exercise=TRUE}

```

```{r c8c-check}
grade_result(
  pass_if(~identical(.result, 0.87))
)
```

<div id="c8c-hint">
**Hint:** From the law of conditional probability: $P(B|A) = \frac{P(A \cap B)}{P(A)}$.
</div> 


## Bayes Rule

- Only one in 1,000 adults has a rare disease for which a screening test has been developed. The test is such that when an individual has the disease, a positive test will result 98% of the time, whereas for an individual without the disease, it will show a positive test result only 1% of the time (false positivity rate). If a randomly selected individual is tested and the result is positive, what is the probability that the individual has the disease? (rounded up to 3 dp)

```{r bayes-setup, include=FALSE}
pD <- 0.001 # pr. has the disease
pnD <- 1-pD # pr. no disease

pTgD <- 0.98  # test sensitivity 
pTgnD <- 0.01 # false positive rate

# pr. has the disease given tests positive
pDgT <- (pTgD*pD)/(pTgD*pD + pTgnD*pnD); round(pDgT,3)
```


```{r bayes, exercise=TRUE}

```

```{r bayes-check}
grade_result(
  pass_if(~identical(.result, 0.089))
)
```

<div id="bayes-hint">
**Hint:** $D=$ event that has the disease, and $P(D) = 1/1000=0.001$, $P(D') = 1 - P(D) = 0.999$. $T=$ event that the person tests positive. Given, $P(T|D) = 0.98$ and $P(T|D')=0.01$. Now, use Bayes' rule to compute: $P(D|T) = \frac{P(T|D) \cdot P(D)}{P(T|D) \cdot P(D) + P(T|D') \cdot P(D')}$ 
</div> 


- A taxi was involved in a minor accident. The town has two taxi companies – Blue and Green. The green taxi company owns 85% of the cabs on the road. A witness says that she saw a blue taxi, but independent tests suggest that she makes a correct identification 80% of the time. What is the probability that a green taxi was in fact involved in the accident? (rounded up to 3 dp)

```{r bayes1-setup, include=FALSE}
pBlue <- 0.15
pGreen <- 0.85
# C = claim that the perpetrator is a Blue taxi
pCgBlue <- 0.80    # correct id
pCgGreen <- 0.20   # incorrect id (false positive)
# prob that a green taxi was the perpetrator
prob <- (pCgGreen*pGreen)/(pCgGreen*pGreen + pCgBlue*pBlue); round(prob, 3)
```

```{r bayes1, exercise=TRUE}

```

```{r bayes1-check}
grade_result(
  pass_if(~identical(.result, 0.586))
)
```

<div id="bayes1-hint">
**Hint:** $P(Blue)=0.15$ (prob of a blue taxi); $P(Green)=0.85$ (prob of a green taxi); $C=$ witness claim that the taxi is blue. Draw a prob. tree. Find $P(Green|C) = \frac{P(C|Green)\cdot P(Green)}{P(C|Green)\cdot P(Green) + P(C|Blue) \cdot P(Blue)}$.
</div> 
