---
title: "Introduction to R"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: readable
    highlight: pygments
runtime: shiny_prerendered
description: >
  A brief introduction to computing in R.
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction 

*Here is a brief introduction to working with the R programming language.*

**R as a super calculator**

If you want to add two numbers you can run the command below:

```{r two-plus-two, exercise=TRUE}
1 + 2
```

Test if you can subtract two numbers in R, say `5` from `8`:

```{r subtract, exercise=TRUE}

```

```{r subtract-solution}
8-5
```

What if we wanted to multiply `5` and `8`?

```{r mult, exercise=TRUE}

```

```{r mult-solution}
5*8
```

**Variable Assignment**

We can assign values to a variable using the assignment operator `<-` or `=`. We can then work with the variables with assigned values.

```{r assign, exercise=TRUE}
x <- 12 # assign the value of 12 to the variable x
y <- 5  # assign the value of 5 to the variable y

product_xy <- x * y # product of x and y

product_xy
```

*Note: anything that starts with a `#` sign is a comment and R will not execute it*

**Installing Packages**

R has extensive user-written packages. If the packages are available on the Comprehensive R Archive Network (CRAN), you can install them by using the command `install.packages("<package-name>")` in R console or `Tools>Install Packages` and search for the package in RStudio.

See [here](https://rstudio-education.github.io/hopr/packages2.html) for details.

**Help about a function**

In the R Console you can type `help(<query>)` or `?<query>` to read its documentation. For example, if you want to know about the function `sum` you can type `help(sum)` of `?sum` and its documentation will show up.

If you are unsure about the name of a function, you can type `apropos("<query>")` to display a list of the functions. For example, `apropos("sum")` shows the result from a fuzzy search with our query.

**Writing your own function**

You can write your own functions in R.

Suppose we want to write a function that multiplies `x` and `y`. We can give the function a name, say `product`. 

```{r func-product, exercise=TRUE}
product <- function(x, y) {
  # function to calculate the product of x and y
  res <- x * y
  print(paste(x, "multiplied by", y, "is", res))
}

product(5, 4)           # calling the function product

product(x = 5, y = 4)   # use named arguments

```

We multiplied `5` with `4`. 

You can change the numbers and re-run the code in the above chunk.
